\sysfull

### 2017-08-07 paper/background.tex
Background section revised (yaay)

Energy harvesting devices operate using energy extracted from their environment, from sources such as radio waves (RF) and solar energy. Energy-harvesting devices need not use a tethered power supply or battery and instead, typically collect energy into a (super-)capacitor, operate when sufficient energy has accumulated, and upon depleting the stored energy, turn off and wait to collect more energy.

@@

Software has different behavior when executed intermittently than when executed with continuous power.  Each power failure clears volatile memory (registers, stack, and global variables). Non-volatile memory (e.g., FRAM) persists across power failures. On a power failure, control flows back to a prior point in the execution. By default, after a power failure control flows to the beginning of {\tt main()}. Early work on intermittent computing preserve intermittent progress by periodically capturing and restoring to a non-volatile checkpoint of the volatile execution context (i.e., registers and stack)~\cite{mementos,quickrecall}, sometimes using hardware support~\cite{mementos,mottola2017harvos,hibernusplusplus,hibernus,idetic}.

@@

%\TODO{This sequential/modular split seems pretty odd to me. I think we can just talk about tasks in the sense of ratchet, alpaca, and chain. Anything with static (programmer-written or compiler inserted) boundaries applies}

@@

\begin{table}
	\centering
	\footnotesize
	\begin{tabular}{l|l}
		Model & Data Copied to/from NVRAM\\
		\hline
		Mementos~\cite{mementos}                             & Reg. + Stack     \\
		DINO~\cite{dino}                                     & Reg. + WAR vars. \\
		Chain~\cite{chain}                                   & PC   + Channel data\\
		Alpaca~\cite{alpaca}                                 & PC   + WAR vars. \\
		Ratchet~\cite{ratchet}, Clank~\cite{hicks_isca_2017} & Reg. (requires NV main mem.)             \\
	\end{tabular}
	\caption{Memory consistency enforcement overheads of various intermittent execution models. Reg. includes the entire register file, PC is the program counter, Channel data are variables explicitly task-shared by the programmer, and WAR vars. are variables involved in WAR dependences.}
	\label{table:chechpoint_comparison}
\end{table}

@@

\paragraph{Memory Consistency Enforcement Overhead}

Intermittent execution models incur an overhead because they must checkpoint
data~\cite{dino,ratchet,quickrecall,mementos}, manage channels~\cite{chain}, or
privatize and commit data~\cite{alpaca}.
%
Table~\ref{table:chechpoint_comparison} compares the overheads for several
recent intermittent execution models (generalizing discussion from~\cite[Sec.
2.4]{alpaca}.)  The mechanism responsible for the overhead of each model
varies, but all the bulk of overhead in all models is a manipulating
non-volatile memory.  Checkpointing moves volatile memory and registers to and
from non-volatile memory, channel accesses directly manipulate non-volatile
data~\cite{chain}, privatization copies data from and commits data to
non-volatile memory~\cite{alpaca}, and idempotence solutions~\cite{ratchet}
require all of memory to be non-volatile.

@@

To concretely understand the cost of these manipulations of non-volatile
memory, we directly measured the energy consumption of volatile (SRAM) and
non-volatile (FRAM) memory operations for the
MSP430FR5969~\cite{msp430datasheet} microcontroller. We used EDB~\cite{edb} to
monitor the voltage drop on an enegy buffering capacitor that powers the device
across 1600 read and write accesses to each type of memory, randomizing
locations to avoid caching effects.  Figure~\ref{fig:framEnergy} summarizes
these measurements, illustrating that a non-volatile memory access consumes
around 1.5x as much energy as a volatile memory access in this device.  In an
intermittently-powered device, the increased energy corresponds to decreased
run time between power failures.  The overheads imposed by relying on
non-volatile memory motivate \sys, which uses memory virtualization to operate
primarily using volatile SRAM.

The above result reassures the intuitive observation that a decrease of CP
(Table~\ref{table:chechpoint_comparison}) decreases the energy cost associated
with each state preservation method and reduces operation cost which in most
simplistic terms is defined as $\mathcal{O}(\text{CP})+\mathcal{O}(\text{PL})$,
where PL is the cost of program execution without checkpointing. On the other
hand, this increases task/checkpoint re-execution time. The ideal solution is a
runtime that \emph{adaptively} changes task size during runtime, which however
exposes a second problem.

%\TODO{Brandon left off here - I've cut some bullets below that I did not understand.  I think the key point here is that fixed task size only works for one cap size; inefficient with a larger cap, unimplementable with a smaller one.  There might be an efficiency argument, but it's tough to make because we still require some memory consistency mechanism (in our case paged FRAM), which has an overhead.}
\paragraph{Fixed-size tasks are inflexible and inefficient} 
A programmer must write each task with statically defined boundaries, which
makes statically sized tasks inflexible and non-portable.  The programmer must
ensure statically that a task's energy requirements match the device's energy
storage capacity.  If a programmer writes a program that consumes more energy
than the device can buffer, the task will never be able to complete using the
buffered energy. The consequence of such tasks is a non-terminating program
that repeatedly attempts to re-execute the task of excess energy cost.  If a
programmer writes a program that consumes less energy than the device can
buffer, the system may operate inefficiently. When a task and its successor
both complete , the two tasks were interleaved by a task boundary that
preserves progress and state (i.e., checkpoint or commit).  If the two tasks
were not also interleaved by a power failure, then the computation done to
preserve state was {\em unnecessary}, only incurring an overhead on the tasks'
executions.   
%
Avoiding excessively costly, non-terminating tasks and short, high-overhead
tasks is a programming challenge, given fixed hardware with a fixed energy
buffering capacity.   Such tasks also make it extremely difficult to {\em port}
code from  one device to another. A task that is excessively costly on a device
with a small energy buffer may become a short, high-overhead task  on a device
with a larger energy buffer and vice versa. The key problem is that the programmer
writes tasks statically for an energy buffer of a fixed size.
%
\sys's {\em task coalescing} mechanism, which we describe in
Section~\ref{sec:task_coalescing}, is motivated by the observation that systems
with statically defined tasks may be inflexible and inefficient; \sys addresses
the problem

@@

\sys is designed for the demands of existing and future energy-harvesting
platforms based around general purpose, commodity computing
components~\cite{wisp,msp430datasheet}.  
%
We assume a device with a memory system that has fast, byte-addressable
volatile and non-volatile memory; in particular, our target
platform~\cite{wisp} is equpped with a mixture of SRAM and FRAM.
%
Our implementation leverages hardware support for fast, bulk-copying between
memories via DMA~\cite{msp430datasheet}.
%
We do not require a particular non-volatile memory technology, nor do we
require architectural additions commodity
processors.~\cite{su_date_2017,ratchet,quickrecall,nvp}. 
%
\sys supports I/O behavior similar to recent intermittent execution
models~\cite{alpaca,chain}, allowing safe, synchronous I/O and unsafe,
asynchronous I/O.

>>>>

Energy harvesting devices operate using energy extracted from their environment, from sources such as radio waves and solar energy. Energy-harvesting devices need not use a tethered power supply or battery and instead, typically collect energy into a (super-)capacitor, operate when sufficient energy has accumulated, and upon depleting the stored energy, turn off and wait to collect more energy.

@@

Software has different behavior when executed intermittently than when executed with continuous power. Each power failure clears volatile memory (registers, stack, and global variables). Non-volatile memory (e.g., FRAM) persists across power failures. On a power failure, control flows back to a prior point in the execution. By default, after a power failure control flows to the beginning of {\tt main()}. Early work on intermittent computing preserve intermittent progress by periodically capturing and restoring to a non-volatile checkpoint of the volatile execution context, i.e., registers and stack~\cite{mementos,quickrecall}, sometimes using hardware support~\cite{mementos,mottola2017harvos,hibernusplusplus,hibernus,idetic}.

@@

@@

\begin{table}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|}
		\hline
		Model & Data Copied to/from NVRAM \\
		\hline\hline
		Mementos~\cite{mementos}                             & Reg. + Stack     \\
		DINO~\cite{dino}                                     & Reg. + WAR variables \\
		Chain~\cite{chain}                                   & PC   + Channel data\\
		Alpaca~\cite{alpaca}                                 & PC   + WAR variables \\
		Ratchet~\cite{ratchet}, Clank~\cite{hicks_isca_2017} & Reg. (requires NV main memory) \\
		\hline
	\end{tabular}
	\caption{Memory consistency enforcement overheads of various intermittent execution models. \emph{Reg.} includes the entire register file, \emph{PC} is the program counter, \emph{Channel data} are variables explicitly task-shared by the programmer, and \emph{WAR variables} are variables involved in WAR dependences (\emph{NV}: non-volatile).}
	\label{table:chechpoint_comparison}
\end{table}

@@

\text{Memory Consistency Enforcement Overhead:} Intermittent execution models incur an overhead because they must checkpoint data~\cite{dino,ratchet,quickrecall,mementos}, manage channels~\cite{chain}, or privatize and commit data~\cite{alpaca}. Table~\ref{table:chechpoint_comparison} compares the overheads for several recent intermittent execution models (generalizing discussion from~\cite[Sec. 2.4]{alpaca}.) The mechanism responsible for the overhead of each model varies, but all the bulk of overhead in all models is a manipulating non-volatile memory. Checkpointing moves volatile memory and registers to and from non-volatile memory, channel accesses directly manipulate non-volatile data~\cite{chain}, privatization copies data from and commits data to non-volatile memory~\cite{alpaca}, and idempotence solutions~\cite{ratchet} require all of memory to be non-volatile.

@@

To concretely understand the cost of these manipulations of non-volatile memory, we directly measured the energy consumption of volatile (SRAM) and non-volatile (FRAM) memory operations for the MSP430FR5969~\cite{msp430datasheet} microcontroller. We used EDB~\cite{edb} to monitor the voltage drop on an energy buffering capacitor that powers the device across 1600 read and write accesses to each type of memory, randomizing locations to avoid caching effects. Figure~\ref{fig:framEnergy} summarizes these measurements, illustrating that a non-volatile memory access consumes around 1.5 times as much energy as a volatile memory access in this device. In an intermittently-powered device, the increased energy corresponds to decreased run time between power failures. The overheads imposed by relying on non-volatile memory motivate \sys, which uses memory virtualization to operate primarily using volatile SRAM.

The above result reassures the intuitive observation that a decrease of checkpoint frequency (CP) decreases the energy cost associated with each state preservation method and reduces operation cost which in most simplistic terms is defined as $\mathcal{O}(\text{CP})+\mathcal{O}(\text{PL})$, where PL is the cost of program execution without checkpointing. On the other hand, this increases task/checkpoint re-execution time. The ideal solution is a runtime that \emph{adaptively} changes task size during runtime, which however exposes a second problem.

\textbf{Fixed-size Tasks are Inflexible and Inefficient:} A programmer must write each task with statically defined boundaries, which makes statically sized tasks inflexible and non-portable. The programmer must ensure statically that a task's energy requirements match the device's energy storage capacity. If a programmer writes a program that consumes more energy than the device can buffer, the task will never be able to complete using the buffered energy. The consequence of such tasks is a non-terminating program that repeatedly attempts to re-execute the task of excess energy cost. If a programmer writes a program that consumes less energy than the device can buffer, the system may operate inefficiently. When a task and its successor both complete, the two tasks were interleaved by a task boundary that preserves progress and state (i.e., checkpoint or commit).  If the two tasks were not also interleaved by a power failure, then the computation done to preserve state was {\em unnecessary}, only incurring an overhead on the tasks' executions. Avoiding excessively costly, non-terminating tasks and short, high-overhead tasks is a programming challenge, given fixed hardware with a fixed energy buffering capacity that might be \emph{different} for each transiently-powered device (see Table~\ref{table:capacitor}). Such tasks also make it extremely difficult to {\em port} code from  one device to another. A task that is excessively costly on a device with a small energy buffer may become a short, high-overhead task on a device with a larger energy buffer and vice versa. The key problem is that the programmer writes tasks statically for an energy buffer of a fixed size. \sys's {\em task coalescing} mechanism, which we describe in Section~\ref{sec:task_coalescing}, is motivated by the observation that systems with statically defined tasks may be inflexible and inefficient.

@@

\sys is designed for the demands of existing and future energy-harvesting platforms based around general purpose, commodity computing components~\cite{wisp,msp430datasheet}. We assume a device with a memory system that has fast, byte-addressable volatile and non-volatile memory; in particular, our target platform~\cite{wisp} is equipped with a mixture of SRAM and FRAM. Our implementation leverages hardware support for fast, bulk-copying between memories via DMA~\cite{msp430datasheet}. We do not require a particular non-volatile memory technology, nor do we require architectural additions commodity processors.~\cite{su_date_2017,ratchet,quickrecall,nvp}. \sys supports I/O behavior similar to recent intermittent execution models~\cite{alpaca,chain}, allowing safe, synchronous I/O and unsafe, asynchronous I/O.
